On the Synchronization of Computational RFIDs

### 2017-11-01 appendix_tmc.tex
TMC submission

\documentclass[10pt,journal,compsoc]{IEEEtran}

\ifCLASSOPTIONcompsoc
  % IEEE Computer Society needs nocompress option
  % requires cite.sty v4.0 or later (November 2003)
  \usepackage[nocompress]{cite}
\else
  % normal IEEE
  \usepackage{cite}
\fi

\usepackage{algorithm,setspace}
\usepackage[noend]{algpseudocode} % needed for: algorithm pseudocode
\usepackage{color}
\usepackage{array}
\usepackage{multirow}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{url}
\usepackage{hyperref}
\usepackage{stmaryrd}
\usepackage{soul}
\usepackage{cancel}
\usepackage{listings}
\usepackage{tikz} % needed for: tikz pics
\usetikzlibrary{arrows}
\usetikzlibrary{decorations.pathreplacing}

\definecolor{aliceblue}{rgb}{0.94, 0.97, 1.0}
\definecolor{applegreen}{rgb}{0.55, 0.71, 0.0}
\definecolor{amber}{rgb}{1.0, 0.75, 0.0}

\lstdefinestyle{mystyle}{
	tabsize          = 1,     % tab space width
	numbers          = left,  % display line numbers on the left
	%stepnumber		 = 2,
	numberfirstline=false,
	%framesep         = 1pt,   % expand outward
	framerule        = 0.2pt, % expand outward 
	commentstyle     = \color{applegreen},      % comment color
	keywordstyle     = \color{blue},       % keyword color
	stringstyle      = \color{amber},    % string color
	backgroundcolor  = \color{aliceblue}, % backgroundcolor color
	showstringspaces = false,              % do not mark spaces in strings
	otherkeywords	 ={signal,inos_task,uint16_t},
	language = C,
	%numbers = none, 
	escapechar = !,
	basicstyle = \ttfamily\bfseries\scriptsize, 
	xleftmargin=0.5ex,
	linewidth = .75\linewidth,
	basicstyle=\tiny,
}

\newcommand{\expected}[1]{E\!\left\llbracket #1 \right\rrbracket}
\newcommand{\revised}[1]{\textcolor{blue}{#1}}
\newcommand{\TODO}[1]{\textcolor{blue}{\textbf{TODO: }}\textcolor{red}{#1}}
\newtheorem{fact}{Fact}

% text revision commands
\newcommand{\add}[1]{\textcolor{blue}{#1}}
\newcommand{\remove}[1]{\textcolor{red}{\st{#1}}}
\newcommand{\replace}[2]{\remove{#1} \add{#2}}

\begin{document}

\title{On the Synchronization of Computational RFIDs}

\author{Kas{\i}m~Sinan~Y{\i}ld{\i}r{\i}m, 
	Henko~Aantjes, 
	Przemys{\l}a~Pawe{\l}czak~\IEEEmembership{Member,~IEEE,} 
	and~Amjad~Yousef~Majid
	%
	\IEEEcompsocitemizethanks{
		\IEEEcompsocthanksitem A preliminary version of this article was presented at the HLPC 2016 Workshop, Atlanta, GA, USA, April 2, 2016~\cite{hlpc2016}. \protect\\
		\IEEEcompsocthanksitem The authors are with the Embedded Software Group, Delft University of Technology, Mekelweg 4, 2628 CD Delft, The Netherlands (e-mail: \{k.s.yildirim, p.pawelczak, a.y.majid\}@tudelft.nl, h.aantjes@student.tudelft.nl).
	}
%	\thanks{Manuscript received April 19, 2005; revised August 26, 2015.}
}

\IEEEtitleabstractindextext{%
\begin{abstract}
Battery-free computational RFID platforms, such as WISP (Wireless Identification and Sensing Platform), are intermittently-powered devices designed for replacing existing sensor networks. Accordingly, synchronization appears as one of the  crucial building blocks for collaborative and coordinated actions in these platforms. However, intermittent power leads to frequent loss of computational state and short-term clock frequency instability that makes synchronization challenging. In this article, we introduce WISP-Sync protocol that provides synchronization among WISP tags in the communication range of an RFID reader. WISP-Sync overcomes the aforementioned challenges by employing a Proportional-Integral (PI) controller-inspired algorithm which (i) is adaptive---reactive to short-term clock instabilities; (ii) requires only a few computation steps---suitable for limited harvested energy; and (iii) keeps a few variables to hold the synchronization state---minimum overhead to recover from power interrupts.	Evaluations in our testbed showed that WISP-Sync ensured an average synchronization error of approximately 1\,ms among the tags with an average energy overhead of 1.85\,$\mu$J per synchronization round.
\end{abstract}

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
Time Synchronization, Computational RFIDs, Wireless Identification and Sensing Platform (WISP).
\end{IEEEkeywords}}

\maketitle

\IEEEdisplaynontitleabstractindextext

\IEEEpeerreviewmaketitle

\IEEEraisesectionheading{\section{Introduction}\label{sec:Intro}}

\IEEEPARstart{L}{ow-power} wireless embedded systems consist of tiny, low-cost and spatially separated computers that communicate with each other by exchanging radio packets. Powering these small-scale embedded systems, e.g. wireless sensor networks (WSNs), is still a crucial problem~\cite{emergence}. Replacing or recharging their batteries is impractical and inhibits long-term operation. Moreover, batteries increase the size and cost of the hardware. Fortunately, the energy efficiency of these systems has improved considerably such that their power requirements are in the order of a few $\mu$W \cite{booksmith2013}. Furthermore, recent advancements in microelectronics technology enabled harvesting power from radio frequency (RF) sources that is sufficient to operate low-power embedded systems in practice~\cite{60ghz,emergence,bi:commag:2015,xie_wcom_2013,garnica_procieee_2013}. Nowadays, the growth of the RF-powered computing paradigm is bringing new research opportunities and challenges~\cite{emergence}, leading to a promising class of low-power embedded systems, the so-called Intermittently Powered Devices (IPDs).

By taking existing RFID (Radio Frequency IDentification) technology as a foundation, computational RFIDs (CRFIDs) are emerging IPDs that allow sensing, computation and communication without batteries---replacing existing battery-powered sensor networks~\cite{booksmith2013}. CRFIDs are equipped with a backscatter radio~\cite[Chapter 4]{rfid_handbook} composed of a simple circuitry that modulates the carrier wave generated by a reader to transmit information. This allows communication to come almost for free. This is a fundamental difference from sensor networks where the transceiver circuit is the most energy-hungry component. In the CRFID domain, the bottleneck in terms of power consumption has shifted from communication to computation and sensing~\cite{ekhonet}. A typical example of CRFIDs is the WISP (Wireless Identification and Sensing Platform)~\cite{wisp}. Commercial RFID readers implementing the EPC Gen2 standard~\cite{epc_gen2} are used to power WISP tags. Apart from low data rate sensing~\cite{neuralWISP}, these maintenance-free devices are evolving to support high data rate and more complex sensing applications such as continuous sensor-data streaming, e.g. capture and transfer of images via battery-free cameras, i.e. WISPCam~\cite{wispcam_2015,cameralocalize_2015}.

\subsection{The Need for CRFID Network Synchronization}

As of now, battery-less networks are in the form of one-hop architecture in which the CRFID tags communicate only with the RFID reader~\cite{wisp,booksmith2013,wisent:2016}. However, recent advancements in the wireless communication enable tag-to-tag communication, e.g.~\cite{liu2013ambient}. We anticipate that these multi-hop networks will demand their own implementation of basic sensor network building blocks~\cite{levis:tinyos} with the aim of replacing the existing battery-powered sensing and data collection applications with their battery-less counterparts. In turn, we conjecture, that in this new battery-less sensing architecture the synchronization will be a fundamental building block that allows, for instance, temporal ordering of the collected data in order to make inferences at the CRFID tag side. Therefore, establishing a common time notion among the tags, rather than synchronizing RFID readers only, will be a crucial service.

\subsection{Differences From Battery-Powered WSNs }
\label{sec:Motivation}

However, the characteristics of CRFID systems expose fundamentally different challenges than battery-powered sensor networks to implement the synchronization service. The reason is mainly twofold:

\begin{itemize}
	\item \textbf{Challenge I:} CRFID systems should perform computations efficiently using \emph{marginal} and intermittent RF power that leads to frequent loss of computational state, e.g. when the RFID reader moves away from the CRFID tag \cite{ransford:2008}. Therefore, computations pertaining to time synchronization have an extremely short time budget, e.g. should be completed with the other computation tasks within 10--100\,ms in WISP~\cite{Ransford:2011:Mementos}. On the contrary, the synchronization approaches in WSNs, e.g. the de facto protocol \cite{Maroti2004}, keep communication infrequent by pushing computation instead, so that the battery power is used efficiently in order to extend the life-time of the network up to several years.

	\item \textbf{Challenge II:} The continuously varying voltage supply introduces severe hardware instability, e.g. varying oscillator frequencies in short-term affects the stability of the clocks and degrades the accuracy sensing. Therefore, computations pertaining to time synchronization should be reactive against very-frequent voltage effects; e.g. during super-capacitor discharge. On the contrary, in WSNs the input voltage is constant and the temperature is the main factor affecting the clock frequencies~\cite{schmid2010interaction}. Temperature compensation techniques, e.g. \cite{Schmid:2009}, are not applicable to compensate for the voltage instabilities in CRFIDs due to their computational complexity.

\end{itemize}

\subsection{Contributions}

The focus of this article is to answer the question of \emph{how to design a building block that synchronizes intermittently-powered CRFID systems?} To this end, we investigate the WISP platform and provide initial observations and limitations pertaining to the synchronization of these devices. In particular, the main contributions and findings of 
this article are:
%
\begin{itemize}
	\item To the best of our knowledge, this is the first study that focuses on the synchronization of tags in CRFID systems. Even though there are studies on the synchronization of multiple RFID readers, e.g.~\cite{reader_sync}, we are unaware of any study that provides synchronization of individual tags in the communication range of an RFID reader. 
	\item By addressing the aforementioned challenges, we design and implement \emph{the first synchronization primitive} for the tags, namely \emph{WISP-Sync}, inspired by the Proportional-Integral (PI) controllers~\cite{pi2015,yildirim2017adaptive}. WISP-Sync (i) 
	requires only a few computation steps to run efficiently under limited harvested energy and keeps a few variables to hold the synchronization state to recover from power interruptions with minimum overhead---\emph{addressing Challenge I}; (ii) is adaptive to react to clock instabilities in a fast manner---\emph{addressing Challenge II}.
	%
	\item We provide theoretical analysis to prove that this primitive establishes synchronization and to reveal the factors affecting its synchronization performance.
	%
	\item Evaluations in our testbed show that an average synchronization error of approximately 1\,ms can be ensured among tags with an energy overhead of 1.85\,$\mu$J per synchronization round using this  primitive.
\end{itemize}
%
We note that WISP-side implementations throughout the article are done using C and assembly language. All source codes and scripts used to generate the results in the article (including parsing, post-processing and measurement results) are available via~\href{https://goo.gl/RuWIZ2}{https://goo.gl/RuWIZ2}.

The remainder of this article is organized as follows. In Section~\ref{sec:Related-Work} we present the fundamental challenges of synchronizing IPDs. Section~\ref{sec:Clock_Hardware} presents the hardware, in particular clock system of the WISP platform briefly, while Section~\ref{sec:clock_model} provides the mathematical notation to be used throughout the article. In 
Section~\ref{sec:sender_receiver} we propose a sender-receiver synchronization approach between an RFID reader and a single tag. Sections~\ref{sec:event_based} and~\ref{sec:eval-wispsync} present the WISP-Sync protocol and its evaluation in our testbed, respectively. We also present the synchronization of multiple tags in Section~\ref{sec:multi-wisp}. Finally, we provide our conclusions and future research directions in Section~\ref{sec:Conclusion}.

\section{Synchronizing Battery-Free CRFIDs}
\label{sec:Related-Work}

In this section, we present a brief  state of the art on synchronization issues in conventional WSNs. Then, we emphasize the main challenges of synchronizing wireless battery-free embedded systems by focusing on a particular IPD platform: computational RFIDs.

\subsection{Synchronization in Conventional WSNs}
\label{sec:WSN-Sync}

The instability of the clock hardware, delays during communication among sensor nodes, and software methods to establish synchronization are the main factors affecting the synchronization in conventional WSNs.

\subsubsection{Clock Hardware}

In WSNs, each sensor node is equipped with a built-in clock that is implemented as a counter register clocked by a low-cost external crystal oscillator. At each oscillator pulse, i.e. \emph{tick} of the clock, the counter register is incremented. The duration between two consecutive ticks is the \emph{rate} of the built-in clock. Environmental factors such as temperature, supply voltage and aging of the crystal prevent built-in clocks to generate ticks at the exact speed of real-time, leading to bounded \emph{clock drift}. The prominent environmental factor affecting the frequency of the built-in clocks is the temperature~\cite{Schmid:2009,Schmid:2010}. Moreover, \emph{quantization errors} occur with low-frequency built-in clocks, which prevents precise timing measurements.

\subsubsection{Wireless Communication}

Sensor nodes exchange their clock information periodically to compute a \emph{software clock} that represents the synchronized notion of time. A software clock is composed of an offset and a frequency that hold the value and speed difference between the corresponding built-in clock and the reference time, respectively. The difference between the reference time and the software clock is the \emph{synchronization error}. In WSNs, the synchronization error is affected by several sources of errors. The \emph{transmission delay}, defined as the time that passes between the start of the broadcast and the receipt by the receiver node, is the major error source and composed of deterministic and non-deterministic components~\cite{Maroti2004}. Assigning timestamps at the MAC layer removes the deterministic delay components and improves synchronization accuracy. This obligates the use of transceivers that allow assignment of time information to a radio packet just before transmission and reception, e.g. Chipcon CC2420~\cite{chipcon:cc2420}. Since the transceiver is the most power-hungry circuit, nodes in WSNs should reduce re-synchronization frequency to decrease communication overhead and save power.

\subsubsection{State-of-the-art Synchronization Techniques}

The most common synchronization mechanism is to propagate the time information of a particular \emph{reference node} to let receiver nodes synchronize themselves to the received reference time information. Commonly, \emph{least-squares regression} is employed by several practical synchronization protocols to adjust the offset and the frequency of the software clocks ~\cite{Maroti2004,Lenzen:2015,TATS:2016}. On the other hand \emph{iterative} computation methods, e.g. in ~\cite{Yildirim:AVTS:2014,pi2015,yildirim2017adaptive,Yildirim:Gradient:2016}, perform identical computation steps at each re-synchronization round. With iterative methods, sensor nodes establish synchronization after a finite number of rounds in a lightweight manner in terms of computation and memory requirements. There are also \emph{fully distributed} approaches in which sensor nodes interact only with and synchronize to their direct neighbors in a peer-to-peer fashion without the requirement of a dedicated reference node. In these approaches, nodes employ computation methods based on \emph{distributed consensus}~\cite{Sommer2009Gradient,SchenatoFiorentin:2011,Yildirim:External:2014}. Since the prominent factor affecting the clock frequencies in WSNs is the temperature~\cite{schmid2010case}, there are also temperature compensated time synchronization techniques, e.g. \cite{Schmid:2009}. However, such techniques require a calibration phase in order to build a temperature--clock frequency table.

\subsection{Fundamental Challenges of Synchronizing CRFIDs}
\label{sec:CRFIDs}

We now delve into the synchronization characteristics of WISP, the de facto CRFID platform. The main aspects pertaining to synchronization in the WISP platform can be stated as follows:

\subsubsection{Single-hop reader-tag architecture} WISP tags are deployed inside the communication range of an RFID reader and they can communicate only with the reader using backscatter communication. Therefore, the RFID reader itself is the natural reference device to establish synchronization among the WISPs, promoting \emph{reader-tag synchronization}. Since WISP tags are unable to communicate with their neighboring nodes directly using backscatter communication as of now, \emph{tag-tag synchronization} is impossible.

\begin{figure}
	\centering
	\includegraphics[width=0.85\columnwidth]{figures/clock_frequency_dc_rf.eps}
	\caption{\label{fig:frequency_dc_rf} In our testbed, see Section~\ref{sec:testbed} for the details, we programmed a single WISP tag so that it toggles one of its output port whenever its 32~KHz built-in timer progresses 32000 clock ticks. In the ideal case where the crystal oscillator of the tag does not drift, the tag would toggle its output port at each 1 second, i.e. with a frequency of 1\,Hz. During our experiment under a stable environmental temperature, we connected the output port of the tag to a logic analyzer and measured the frequency of the toggle event. We plotted our measurements when the tag is powered by the Flash Emulation Tool (FET) with a stable DC power and when the tag is powered by the RFID Reader at a distance of 5~cm. We measured the mean and standard deviation of the frequency in the DC case as 1.1653 and 0.00028553, and these values for the RF case were 1.1793 and 0.00026049, respectively. }
\end{figure}

\subsubsection{Continuously varying voltage level} In WSNs, the battery level decreases gradually that allows stable voltage levels in the short term. On the contrary, on the WISP platform, the fluctuating input voltage prevents short-term stability of the clock hardware and introduces significant drift. Hence, the prominent factor affecting the frequency of the crystal oscillator is the varying voltage level, see Fig.~\ref{fig:frequency_dc_rf}.

\subsubsection{Frequent loss of synchronization state} Contrary to sensor nodes, WISP tags frequently ``die'' due to power loss and they need to save the synchronization state, e.g. the clock offset and relative frequency, into the non-volatile memory to recover when they harvested sufficient energy to start up. However, saving computational state is also an energy consuming task~\cite{blisp}.

\subsubsection{Computation and memory overhead sensitivity} The classical motto of WSNs, ``compute instead of communicate whenever possible''~\cite[p. 44]{karl2007protocols}, is no longer valid for the WISP platform since backscatter communication comes almost for free~\cite{ekhonet}. Due to the intermittent power, lightweight methods in terms of computation and memory are required for the synchronization. Since methods like least-squares regression are computationally heavy and require considerable amount of memory~\cite{pi2015} they should be avoided.

\subsubsection{Limitations of the EPC Gen 2 standard} The WISP firmware implements the EPC Gen 2 standard~\cite{epc_gen2} which increases the compatibility with the existing RFID systems. However, the standard introduces limitations, e.g. currently it does not assign timestamps to the radio packets, which is a fundamental requirement to establish synchronization. Moreover, communication delays between the reader and tag are quite dependent on the implementation of this standard by RFID readers. Unfortunately, these issues lead to less accurate synchronization as compared to existing WSN solutions, as shown by our measurements presented in the following sections. Consequently, WSN synchronization protocols that can guarantee $\mu s$-level synchronization accuracy, e.g.~\cite{Maroti2004}, are hard to achieve in CRFID domain.

\section{WISP Hardware and CRFID Networks: A Brief Look}
\label{sec:Clock_Hardware}

In this section we present brief information about the WISP hardware, in particular its clock system. For the sake of clarity, we 
also give details about a typical CRFID network, which we used to implement and evaluate our synchronization methods in this article.

\subsection{Low Power Operation Modes and The Clock System}
\label{subsec:clock_system}

The WISP 5.0 platform comes with the MSP430FR5969 \cite{msp430fr5969_data_sheet} microcontroller (later called as MSP430 in this article) with FRAM non-volatile memory. The MSP430 clock system includes (i) a 32\,kHz external crystal oscillator, (ii) an internal very-low-power low-frequency oscillator, and (iii) an integrated internal digitally controlled oscillator (DCO). The five built-in 16-bit timers (TA0--TA3, TB0) in this system can be clocked with the auxiliary clock (ACLK) signal that is sourced from the relatively stable external 32\,kHz oscillator. The MSP430 has one active mode and seven software selectable low-power operation modes. Low Power operation Mode 3 (LPM3) is the standby mode where CPU, FRAM and high-frequency peripherals are off; the external 32\,kHz oscillator and the ACKL signal are active. In our target implementations, we configured TB0 to be clocked  with ACKL so that it has 16-bit 32\,kHz precision and runs continuously in LPM3 mode.

\subsection{\label{sec:testbed}An Experimental CRFID System}

\begin{figure}
	\centering
	\begin{tikzpicture}
	\node[anchor=south west, inner sep=0] at 
	(0,0){\includegraphics[width=\columnwidth]{figures/testbed_resized.jpg}};

	\node [fill= white,circle] at (7,2) { };
	\node [color=black ] at (7,2) { 1 };

	\node [fill= white,circle] at (8,4) {  };
	\node [color=black ] at (8,4) { 2 };

	\node [fill= white,circle] at (4,2) {  };
	\node [color=black ] at (4,2) { 3 };

	\node [fill= white,circle] at (5.1,2) {  };
	\node [color=black ] at (5.1,2) { 4 };

	\node [fill= white,circle] at (6.5,4.5) {  };
	\node [color=black ] at (6.5,4.5) { 5 };

	\end{tikzpicture}
	\caption{\label{fig:crfid_testbed} A single-hop CRFID testbed: (1) an Impinj Speedway R1000 RFID reader, (2) a single WISP 5.1 tag, (3) a host computer, (4) FET programmer and (5) a Laird S9028PCR antenna. The same testbed setup has been used previously in~\cite{wisent:2016} by us to evaluate downstream data transmission in CRFIDs.}
\end{figure}

As presented in Fig.~\ref{fig:crfid_testbed}, a typical CRFID system is composed of an RFID reader, a single WISP tag placed inside the communication range of this reader and a host computer to control this reader. In our testbed, we used a 915\,MHz Impinj Speedway R1000 RFID reader with firmware version 3.2.4 connected to a Laird S9028PCR 8.5\,dBic gain antenna. We placed the WISP tag at the line-of-sight from the reader antenna. For the host-reader control operations, we used sllurp~\cite{sllrp_github}, a LLRP (Low-Level Reader Protocol)~\cite{llrp}  control library written in Python.  To program the WISP tag, we used MSP430 Flash Emulation Tool (FET), in combination with TI Code Composer Studio (CCS), attached to the host.  All of our experimental evaluation presented in the next sections were conducted using this setup at a university office with human presence.

\subsection{\label{sec:sniffer}A Sniffer for the CRFID System}

Communication delays are one of the main sources of errors in synchronization~\cite{Maroti2004} and they should be kept as small as possible. In order to characterize the communication delays between the RFID reader and the WISP tags during backscatter communication and understand the effect of these delays on the synchronization error, we used USRP 210 software-defined radio, another Laird antenna placed at 50\,cm from the tag and the GNU Radio toolkit~\cite{gnuradio} with the custom software to sniff the radio packets. The sniffer was presented first in~\cite{stork:2017} and we refer the reader to this study for more details. Implementing sniffing mechanism allowed us to store the timings of the transmission and reception of the EPC Gen 2 packets, which we later processed offline to measure the actual transmission delays. We will present these measurements in the next sections. 
\section{System Model, Notation and Preliminaries}
\label{sec:clock_model}

In this section, we provide a mathematical framework that models the built-in clocks and in particular the clock drifts of the WISP tags and the RFID reader. We will use these models to prove the correctness of our synchronization algorithms and to derive mathematical expressions that represent their synchronization performance.
\begin{figure*}
	\centering
	\includegraphics[width=1.4\columnwidth] {figures/relative_clock_frequency.eps}
	\caption{\label{fig:frequency}The frequencies of the toggle events (obtained using the same steps presented in Fig.\ref{fig:frequency_dc_rf}) of two RF-powered WISP tags, placed at 5 cm distance from the RFID reader. On the left the individual event frequencies ($f_1$ and $f_2$) are presented, and their relative clock frequency ($\frac{f_1}{f_2}-1$) is presented on the right. We measured the mean and standard deviation of $f_1$ as 1.1793 and 0.00026049 and those of $f_2$ as 1.1705 and  0.00068239, respectively.}
\end{figure*}
In this framework, we denote a single WISP tag by $w$ and the RFID reader by $r$. The Timer B0 mentioned in Section~\ref{subsec:clock_system} can be considered as the built-in clock, i.e. \emph{local clock}, of the tag $w$, 
which we denote by $C_{w}$. The value of $C_{w}$ at time $t$ can be modeled as
%
\begin{eqnarray}
C_{w}(t) & \triangleq & C_w(t_0)+ \int_{t_0}^{t}f_{w}(u)du, 
\label{eq:local_clock}
\end{eqnarray}
%
where $t_0$ represents the time at which tag $w$ is powered on and  $f_{w}(u)$ represents the \emph{instantaneous oscillator frequency} of $C_{w}$ at time $u$. Since the frequency of the oscillators are affected by environmental factors such as temperature and voltage level, they are not stable~\cite{Lenzen2009Optimal} and $f_{w}$ is not a constant but it is \emph{time varying}, see Fig.~\ref{fig:frequency}. From \eqref{eq:local_clock}, it can be noticed that $C_w$ is \emph{unitless}. Indeed, it can be seen as a counter of oscillation events that are produced at a frequency of $f_w$.

The RFID reader $r$ can be considered as a natural time reference for the tag $w$. We denote the local clock of the RFID reader at time $t$ by $C_{r}(t)$ and its frequency by $f_r(t)$. By collecting (time) information from the RFID reader, the tag $w$ runs a \emph{clock synchronization algorithm} to calculate a \emph{software clock}, denoted by $S_{w}(t)$, whose value represents the synchronized notion of time. The objective of the synchronization algorithm is minimizing the \emph{synchronization error} with respect to the reader $r$ at any time $t$, which is defined as 
%
\begin{eqnarray}
\gamma(t)&\triangleq&S_w(t)-C_r(t). \label{eq:gamma_error}
\end{eqnarray}
%
Considering the individual clock frequencies $f_w$ and $f_r$, we denote the \emph{instantaneous relative clock 
	frequency} of the WISP tag 
with respect to the RFID reader at time $t$ as
%
\begin{eqnarray}
f_w^r(t)&\triangleq&\frac{f_w(t)}{f_r(t)}-1. \label{eq:relative_frequency}	
\end{eqnarray}
\section{Sender-Receiver Based Reader-Tag Synchronization in CRFID Systems}
\label{sec:sender_receiver}

In this section we provide initial observations, design and implementation of a \emph{sender-receiver} based synchronization approach that synchronizes a single WISP tag to the RFID reader and we reveal its limitations in current CRFID networks. In sender-receiver based synchronization mechanisms, receiver devices synchronize to the clock of a reference sender device. In order to synchronize itself to the RFID reader with such a mechanism, the WISP tag should obtain several $(C_w(t),C_r(t))$ synchronization points to establish a \emph{relationship} between its local clock $C_w$ and the reader clock $C_r $, represented by its software clock $S_w$. The value $S_w(t)$ will provide an estimate of the reference clock $C_r(t)$ at any time instant $t$.

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/Sender-Receiver.pdf}
	\caption{\label{fig:sender_receiver} The message exchange among the host computer, the RFID reader and the tag for sender-receiver synchronization. The host machine, i.e. a PC,  sends the high level commands to the reader via LLRP \emph{AccessSpec} message and receives the results through an \emph{ASReport}. The RFID reader follows the steps defined in the EPC Gen2 standard: (i) performs a \emph{Handshake} to initialize the communication with the active tag; (ii) performs a \emph{Command Initialization} (Cmd Init) by requesting a random number from the tag (Req\_n) and receiving the random number (newRN); (iii)  performs a \emph{Read} command by sending its request and receiving the timer value. The reader assigns the \emph{FirstSeenTimestamp} to the tag at time $t_1$ and the tag timestamps the command reception event at time $t_2$ with its local clock reading.}
\end{figure}

We explored the EPC Gen2 standard and the LLRP protocol and found out that LLRP assigns a \emph{FirstSeenTimestamp} in UTC (Coordinated Universal Time), which is defined as \emph{``The Reader SHALL set it to the time of the first observation amongst the tag reports that get accumulated in the TagReportData''} \cite[p. 87]{llrp}. From this definition, we assume that this timestamp is assigned by the reader when it receives the EPC during the handshake operation with the corresponding tag, shown in Fig. \ref{fig:sender_receiver} as the timestamp assigned at time $t_1$. Therefore, the \emph{FirstSeenTimestamp} can be considered as $C_r(t_1)$. In order to obtain the corresponding local time $C_w(t_1)$, one strategy is to force the reader to send a special ``synchronization'' command after the handshake so that the tag timestamps the command reception event using its local clock, shown in Fig.\,\ref{fig:sender_receiver} as time $t_2$. As we emphasized in Section~\ref{sec:WSN-Sync}, transmission delays are the main source of errors in time synchronization and they should be kept as deterministic as possible. We denote the \emph{transmission delay} in this case by $\Delta t= t_2-t_1$ and it is desirable to keep the \emph{variation} of $\Delta t$, i.e. \emph{jitter}, as small as possible~\cite{Maroti2004}.

\subsection{Transmission Delay and Jitter Measurements}

\begin{figure}
	\centering
	\includegraphics[width=0.9\columnwidth]{figures/sender_receiver_uncertainty.eps}
	\caption{\label{fig:transmission_delay} The normalized number of occurrences of the transmission delays measured by sniffing the communication between the RFID reader and the WISP tag. We calculated the mean transmission delay and its standard deviation as 1.89\,ms and 0.0164\,ms with a 99\% confidence interval of [1.8874,1.8925] and [0.0148,0.0184], respectively.}
\end{figure}

In order to characterize the transmission delay $\Delta t$, we placed the WISP tag 20 cm away from the antenna using the setup presented in Fig.~\ref{fig:crfid_testbed} and we sniffed the communication between the RFID reader and the WISP tag during the communication scenario of Fig.~\ref{fig:sender_receiver}. Fig.~\ref{fig:transmission_delay} presents a summary of our measurements (with 300 samples gathered). We observed that the transmission delay is distributed with a mean of 1.89\,ms and standard deviation of 0.0164\,ms. Moreover, we observed some outliers, presented at the right hand side of the figure, which we attribute to the EPC Gen2 implementation of the Impinj reader.

\subsection{Collection of Clock Values for Off-line Processing}

After characterizing the variation of the transmission delay, the next step was the collection of the clock values from the RFID Reader and the WISP tag in order to try synchronization mechanisms offline, which allowed to evaluate the performance of sender-receiver synchronization mechanism in CRFID networks. We collected $(C_w(t_2),C_r(t_1))$ pairs for offline processing by controlling the RFID reader to send a \emph{Read} command to the tag and by programming the WISP tag so that it backscatters $C_w(t_2)$ upon receiving this command. The received pairs $(C_w(t_2),C_r(t_1))$ were logged by the host computer.

\subsection{Software Clock Computation Using  Least-Squares}

Following studies~\cite{Maroti2004,Schmid:2010,Lenzen:2015,TATS:2016} that use a linear clock model, we assume a \emph{linear} relationship between $C_r$ and $C_w$ and we model the software clock of the WISP tag as $S_w(C_w(t)) = a + b C_w(t)$ where $S_w$ represents an estimator of the reader's clock $C_r$, $a$ is the \emph{offset} and $b$ is the \emph{relative speed} with respect to the local clock $C_w$. To establish such a relationship, we performed \emph{least-squares regression} offline in MATLAB using the collected pairs in the previous subsection. As in \cite{Maroti2004}, since the WISP tag has limited memory, computation capability and energy, at each step only the most recent $N$ pairs are used to estimate the regression line.

\begin{figure}
	\centering
	\includegraphics[width=0.9\columnwidth]{figures/least_squares.eps}
	\caption{\label{fig:least-squares}Synchronization error by employing least-squares regression on the collected timestamps. We observed a maximum synchronization error of 10 clock ticks between the RFID reader and the WISP tag, leading to 0.32\,ms synchronization accuracy.}
\end{figure}

Formally, let $[C_w(t_k),C_r(t_k)]$ denote the $k$th pair in the log file where $t_k$ denotes the real-time at which 
\emph{FirstSeenTimestamp} has been assigned during the collection of this pair. At each $k$th step, the pairs $\{[C_w(t_k),C_r(t_k)],$\ldots$,[(C_w(t_{k+N-1}),C_r(t_{k+N-1})]\}$ are used to calculate the parameters $a$ and $b$, i.e. the \emph{intercept} and the \emph{slope} of the estimated regression line \cite{book-SheldonRoss}, using \textsf{polyfit} function of MATLAB. We calculated the synchronization error as 
$\gamma(t_{k+N})\triangleq S_w(C_w(t_{k+N}))-C_r(t_{k+N})$
%
that represents the difference between the predicted reference time and the received reference time. In our implementation we used $N=8$ as in~\cite{Maroti2004} and Fig.~\ref{fig:least-squares} presents the synchronization error at each step. We observed a maximum synchronization error of 0.32\,ms in this one-hop network, which is more than one order of magnitude larger than the synchronization performance of the de facto WSN solution \cite{Maroti2004}, which was reported as approximately 10\,$\mu$s.

\subsection{Limitations of the Approach}

In addition to the challenges listed in Section \ref{sec:CRFIDs}, we observed two crucial limitations for the WISP platform, which prevents to build up a sender-receiver synchronization building block:

\subsubsection{Host computer computation} We are unaware of any EPC Gen2 command that will allow to send the \emph{FistSeenTimestamp} to the tag. Hence, even though we were able to collect $(C_w(t),C_r(t))$ pairs for offline processing, it is not possible for the tag to collect $C_r(t)$ and synchronize itself to the reader. Therefore, with this limitation, only a host computer can collect and log the timestamps, calculate the relationship between the clock of the tag and the clock of the reader and send the data that represents this relationship to the tag for synchronization, making this method only of theoretical nature as of now.

\subsubsection{The Lack of broadcast primitive} Since the RFID reader assigns the \emph{FirstSeenTimestamp} for each tag, the synchronization steps in Fig.~\ref{fig:sender_receiver} should be repeated for each tag to obtain synchronization in the communication domain of the reader. Unfortunately, in the current EPC Gen2 and LLRP standard, we were unable to find any mechanism to send a global timestamp that is received by all tags inside the range of the RFID reader to establish synchronization in one step.

\subsubsection{Big and reader dependent transmission delays} As depicted previously, we observed that transmission delays in our testbed were on the order of milliseconds, see Fig.~\ref{fig:least-squares}, whereas they are reported as a few microseconds in conventional WSN platforms. This is one of the crucial limitations since the smaller the transmission delays are, the better the synchronization.

\section{Event-Based Reader-Tag Synchronization In CRFID Systems}\label{sec:event_based}

In this section, we introduce a simple but novel \emph{event-based} reader-tag synchronization inspired by simple PI controllers and provide its advantages over the previous approach. In event-based synchronization, a common event which is observable by all receiver devices simultaneously is generated by the reference device. Upon receiving events generated at regular intervals, receiver devices can predict the occurrence time of future events. In order to synchronize the tag with such a mechanism, the RFID reader does not send explicit timestamp values as in sender-receiver based synchronization but instead generates events at regular intervals. Upon observing these events, the receiver tag adjusts the rate of its software clock so that it predicts the occurrence of the next event precisely. 
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/Event-Based.pdf}
	\caption{\label{fig:event_based}Event-Based synchronization steps: The tag timestamps successive \emph{BlockWrite} events and adjusts its software clock.}
\end{figure}
We explored the EPC Gen2 standard to see how to generate events at regular intervals and realized that the \emph{BlockWrite} operation allows this feature.\footnote{It is possible to exploit modulation properties of signals from the reader to tags (such as bit timing in FM0, i.e. Miller 1, modulation) in order to get the timing measurements of the regular patterns in the signal. However, this requires using a high-precision clock to obtain microsecond-precision timing measurements. In this study, we used the higher-level properties of the EPC Gen2 standard, rather than exploiting its low-level signal properties. By means of this, we could  use the external 32\,KHz clock that runs even in low-power standby mode (see Section \ref{sec:Clock_Hardware}) to obtain timing measurements.} Fig.~\ref{fig:event_based} presents the steps of event-based synchronization. During the command phase, the EPC Gen2 standard allows a maximum of eight successive \emph{BlockWrite} operations. It is desirable to use the first and the last \emph{BlockWrite} events for synchronization since it is better to compensate frequency differences observed in longer time intervals to adjust the software clock.

\subsection{Event Period Measurements}
\label{subsec:event_period_uncertainty}

>>>>

\newpage\null\thispagestyle{empty}\newpage
\appendices
\section{Analysis of Algorithm~\ref{alg:sync}}
